---
title: "EBNM with Matern"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ebnm_matern}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning=FALSE, message=FALSE}
library(EBMFSmooth)
library(flashier)
library(INLA)
library(Matrix)
library(tidyverse)
library(gridExtra)
```

## A Factor Analysis Model with Spatially-varying Factors

Consider a matrix $Y_{ij}$ that denotes the gene expression collected for gene $i$ at location $s_j \in \mathbb{R}^2$.

We assume there are $n = 50$ genes, each collected $P = 100$ locations. 

Simulate the locations and expression modules:

```{r}
set.seed(1234)
P <- (10)^2
locations <- expand.grid(seq(0,1,length.out = sqrt(P)), seq(0,1,length.out = sqrt(P)))
colnames(locations) <- c("x", "y")
f <- function(x, y, center_x, center_y, scale) {
  exp(-((x - center_x) ^ 2 + (y - center_y) ^ 2) / scale)
}
f1 <- f(locations[,1], locations[,2], center_x = 0.5, center_y = 0.4, scale = 0.2)
f2 <- f(locations[,1], locations[,2], center_x = 0.2, center_y = 0.7, scale = 0.2)
```


Simulate the intensity in the expression modules:

```{r, fig.height = 6, fig.width = 6}
n <- 50
l1 <- rexp(n, rate = 1) * rbinom(n, size = 1, prob = 0.4)
l2 <- rexp(n, rate = 1) * rbinom(n, size = 1, prob = 0.7)

sigma_E <- 0.1
Y <- matrix(0, nrow = n, ncol = P)
Y <- l1 %*% t(f1) + l2 %*% t(f2) + rnorm(n * P, mean = 0, sd = sigma_E)
heatmap(Y, 
        scale = "row",    # Normalize the gene expression data by row
        Colv = NA,        # Don't cluster columns (genes)
        Rowv = NA,        # Don't cluster rows (time points)
        col = colorRampPalette(c("blue", "white", "red"))(50))  # Color gradient
```


## Implementation of EBMF through Flashier

Now, implement the EBMF through `flashier`:

```{r}
ebnm_Matern <- EBMFSmooth::ebnm_Matern_generator(cbind(locations$x, locations$y))
ebnm_list <- list()
ebnm_list[[1]] <- flashier::flash_ebnm(
    prior_family = "point_exponential"
)
ebnm_list[[2]] <- ebnm_Matern
```

```{r}
# record the runtime:
system.time({
  f.mod = flashier::flash(data = (Y), S = (sigma_E), 
                ebnm_fn = ebnm_list,
                greedy_Kmax = 2, var_type = 0,
                backfit = T, 
                verbose = 3)
})
```

Let's take a look at the fitted result:

```{r, fig.height = 6, fig.width = 8}
l1_norm <- l1/sqrt(sum(l1^2))
l2_norm <- l2/sqrt(sum(l2^2))
f1_norm <- f1*sqrt(sum(l1^2))
f2_norm <- f2*sqrt(sum(l2^2))

L_gp_norm1 <- (f.mod$L_pm[,1]) / sqrt(sum(f.mod$L_pm[,1]^2))
L_gp_norm2 <- (f.mod$L_pm[,2]) / sqrt(sum(f.mod$L_pm[,2]^2))
F_gp_norm1 <- (f.mod$F_pm[,1]) * sqrt(sum(f.mod$L_pm[,1]^2)) 
F_gp_norm2 <- (f.mod$F_pm[,2]) * sqrt(sum(f.mod$L_pm[,2]^2))

par(mfrow = c(2,2))
plot(L_gp_norm1 ~ c(1:n), type = "l", col = "red", lty = "dashed",
     ylim = c(min(l1_norm),max(l1_norm)), xlab = "index", ylab = "loadings")
lines(l1_norm ~ c(1:n), type = "l", xlab = "index", ylab = "loadings", lty = "solid")
plot(L_gp_norm2 ~ c(1:n), type = "l", col = "red", lty = "dashed",
     ylim = c(min(l2_norm),max(l2_norm)), xlab = "index", ylab = "loadings")
lines(l2_norm ~ c(1:n), col = "black", lty = "solid")
plot(F_gp_norm1 ~ c(1:P), type = "l", col = "red", lty = "dashed",
     ylim = c(min(f1_norm),max(f1_norm)), xlab = "index", ylab = "factors")
lines(f1_norm ~ c(1:P), type = "o", ylab = "factors", lty = "solid")
plot(F_gp_norm2 ~ c(1:P), type = "l", col = "red", lty = "dashed",
     ylim = c(min(f2_norm),max(f2_norm)), xlab = "index", ylab = "factors")
lines(f2_norm ~ c(1:P), col = "black", lty = "solid", type = "o")
par(mfrow = c(1,1))

# MSE: Normalized Factor
mse_f <- (mean((f1_norm - F_gp_norm1)^2) + mean((f2_norm - F_gp_norm2)^2)) / 2
mse_f
# MSE: Normalized Loading
mse_l <- (mean((l1_norm - L_gp_norm1)^2) + mean((l2_norm - L_gp_norm2)^2)) / 2
mse_l
```


Plot the true spatial surface:

```{r fig.height = 6, fig.width = 8, warning=FALSE, message=FALSE}
grid_res <- 200
fine_locations <- expand.grid(
  x = seq(0, 1, length.out = grid_res),
  y = seq(0, 1, length.out = grid_res)
)

# Evaluate f1 and f2 on the grid
fine_locations$f1 <- f(fine_locations[,1], fine_locations[,2], center_x = 0.5, center_y = 0.4, scale = 0.2)
fine_locations$f2 <- f(fine_locations[,1], fine_locations[,2], center_x = 0.2, center_y = 0.7, scale = 0.2)
fine_locations$f1 <- fine_locations$f1 /max(fine_locations$f1)
fine_locations$f2 <- fine_locations$f2 /max(fine_locations$f2)

# Create the heatplots
plot_f1 <- ggplot(fine_locations, aes(x = x, y = y, fill = f1)) +
  geom_tile() +
  scale_fill_gradientn(colors = terrain.colors(10)) +
  labs(title = "Heatmap of f1", x = "X Coordinate", y = "Y Coordinate", fill = "f1") + 
  # remove the legend
  theme(legend.position = "none") +
  theme_classic()

plot_f2 <- ggplot(fine_locations, aes(x = x, y = y, fill = f2)) +
  geom_tile() +
  scale_fill_gradientn(colors = terrain.colors(10)) +
  labs(title = "Heatmap of f2", x = "X Coordinate", y = "Y Coordinate", fill = "f2") +
  theme_classic()

library(akima)

locations$fitted_f1 <- F_gp_norm1
locations$fitted_f2 <- F_gp_norm2

interp_f1 <- with(locations, interp(x, y, fitted_f1, duplicate = "mean"))
interp_f2 <- with(locations, interp(x, y, fitted_f2, duplicate = "mean"))

interp_f1_df <- data.frame(
  x = rep(interp_f1$x, times = length(interp_f1$y)),
  y = rep(interp_f1$y, each = length(interp_f1$x)),
  fitted_f1 = as.vector(interp_f1$z)
)
interp_f2_df <- data.frame(
  x = rep(interp_f2$x, times = length(interp_f2$y)),
  y = rep(interp_f2$y, each = length(interp_f2$x)),
  fitted_f2 = as.vector(interp_f2$z)
)

interp_f1_df$fitted_f1 <- interp_f1_df$fitted_f1 / max(na.omit(interp_f1_df$fitted_f1))
interp_f2_df$fitted_f2 <- interp_f2_df$fitted_f2 / max(na.omit(interp_f2_df$fitted_f2))

plot_f1_fitted <- ggplot(interp_f1_df, aes(x = x, y = y, fill = fitted_f1)) +
  geom_tile() +
  scale_fill_gradientn(colors = terrain.colors(10)) +
  labs(title = "Heatmap of fitted_f1", x = "X Coordinate", y = "Y Coordinate", fill = "f1") +
  theme_classic()

plot_f2_fitted <- ggplot(interp_f2_df, aes(x = x, y = y, fill = fitted_f2)) +
  geom_tile() +
  scale_fill_gradientn(colors = terrain.colors(10)) +
  labs(title = "Heatmap of fitted_f2", x = "X Coordinate", y = "Y Coordinate", fill = "f2") +
  theme_classic()

# Display the plots in 2x2 grid
grid.arrange(plot_f1, plot_f2, plot_f1_fitted, plot_f2_fitted, ncol = 2)
```
