---
title: "Example of EBMFSmooth"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Iyer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning=FALSE, message=FALSE}
library(EBMFSmooth)
library(fdapace)
library(ggplot2)
```

In this vignette, we will show how to use the `EBMFSmooth` package to perform EBMF analysis on the `Iyer` dataset. 
This dataset is a time-course gene expression dataset with 12 (unequally spaced) time points. 

Before we apply the EBMF analysis, let's load the data and take a look at the first few rows of the dataset.

```{r}
head(EBMFSmooth::Iyer_expression)
```


Let's visualize some example genes from the dataset. 

```{r, fig.width=6, fig.height=6}
plot_expression <- function(i, data_frame) {
  # Check if the index is within bounds
  if (i < 1 || i > nrow(data_frame)) {
    stop("Index out of bounds. Please provide a valid index.")
  }

  # Define actual time points corresponding to the column names
  actual_time <- c(0, 15, 30, 60, 120, 240, 360, 480, 720, 960, 1200, 1440)/60  # Adjust if needed
  if (length(actual_time) != (ncol(data_frame) - 4)) {
    stop("Mismatch between time points and expression data columns.")
  }

  # Extract the expression data for the specified gene
  gene_expression <- data_frame[i, 5:ncol(data_frame)]  # Assuming expression data starts at column 5

  # Create a dataframe for plotting
  plot_data <- data.frame(
    Time = actual_time,
    Expression = as.numeric(gene_expression)
  )

  # Add the gene name for the title
  gene_name <- data_frame$`Gene name`[i]

  # Generate the plot
  ggplot(plot_data, aes(x = Time, y = Expression)) +
    geom_line(color = "blue") +
    geom_point(size = 1, color = "red") +
    labs(
      title = paste("Gene:", gene_name),
      x = "Time (hours)",
      y = "Rel (Log) Expression Level"
    ) +
    theme_minimal() +
    theme(
      title = element_text(size = 6),
      axis.text.x = element_text(angle = 45, hjust = 1, size = 10)
    )
}
p1 <- plot_expression(1, Iyer_expression); p2 <- plot_expression(2, Iyer_expression)
p3 <- plot_expression(3, Iyer_expression); p4 <- plot_expression(4, Iyer_expression)

gridExtra::grid.arrange(p1, p2, p3, p4, nrow = 2)
```


First, let's apply Functional Principal Component Analysis (FPCA) to the gene expression data using the `fdapace` package.

```{r}
expr_matrix <- as.matrix(Iyer_expression[, 5:ncol(Iyer_expression)])
time_points <- c(0, 15, 30, 60, 120, 240, 360, 480, 720, 960, 1200, 1440)/60
```


```{r, fig.width=6, fig.height=6, warning=FALSE, message=FALSE}
N <- nrow(Iyer_expression)
input_pace <- MakeFPCAInputs(tVec = time_points, yVec = (expr_matrix))
FPCA_pace <- FPCA(input_pace$Ly, input_pace$Lt)

plot(FPCA_pace)

score_data_pace <- data.frame(
  PC1 = FPCA_pace$xiEst[, 1],  # Scores for the first principal component
  PC2 = FPCA_pace$xiEst[, 2],  # Scores for the second principal component
  Gene = Iyer_expression$`Gene name`  # Add gene names for labeling (optional)
)
```

There are three eigenfunctions estimated from the FPCA analysis. We could also visualize if there exists any clustering of the genes based on the first two principal components.

```{r, fig.width=6, fig.height=6}
# Scatter plot of the first two principal components
ggplot(score_data_pace, aes(x = PC1, y = PC2)) +
  geom_point(color = "blue", size = 0.5) +
  # geom_text(aes(label = Gene), hjust = 0.5, vjust = -0.5, size = 0.5, check_overlap = TRUE) +
  labs(
    title = "FPCA Score Plot",
    x = "First Principal Component (PC1)",
    y = "Second Principal Component (PC2)"
  ) +
  theme_minimal()
```


By construction of the FPCA, the eigenfunctions are assumed to be orthogonal. 
As a result, each gene is likely to be represented by quite varying weights on the eigenfunctions.
This can be seen from the scatter plot of the first two principal components, where the genes are spread out across the plot.
The interpretation of the FPCA scores therefore becomes more challenging.

As a comparison, let's apply the EBMF analysis to the same dataset.
We assume the factors are smooth functions of time and the loadings are non-negative vectors with a sparsity-inducing prior.

```{r, warning=FALSE, message=FALSE}
expr_matrix <- as.matrix(Iyer_expression[, 5:ncol(Iyer_expression)])
# setup LGP
ebnm_list <- list()
ebnm_list[[1]] <- flashier::flash_ebnm(
  prior_family = "point_exponential"
)
settings <- EBMFSmooth::LGP_setup(t = time_points[-1], p = 2, num_knots = 10, betaprec = 1e-5)
ebnm_list[[2]] <- EBMFSmooth::ebnm_LGP_generator(LGP_setup  = settings)
f.mod <- flashier::flash(
  data = (expr_matrix[, -1]),
  S = NULL,
  ebnm_fn = ebnm_list,
  greedy_Kmax = 3,
  var_type = 1,
  backfit = T,
  verbose = 0
)

L_gp_norm <- sweep(f.mod$L_pm, 2, sqrt(colSums(f.mod$L_pm^2)), FUN = "/")
F_gp_norm <- sweep(f.mod$F_pm, 2, sqrt(colSums(f.mod$L_pm^2)), FUN = "*")
```


```{r fig.width=6, fig.height=6}
matplot(x = time_points[-1], y = F_gp_norm, type = "l", lty = "dashed", xlab = "Time", ylab = "Factors")
```


```{r fig.width=6, fig.height=6}
# Visualization function for stacked bar plot of L_pm
plot_L_pm <- function(L_pm, order = 1) {
  # Validate the 'order' argument
  if (!is.numeric(order) || order < 1 || order > ncol(L_pm)) {
    stop("`order` must be a numeric value between 1 and the number of factors.")
  }
  
  # Sort genes by their intensity in the specified factor
  sorted_indices <- order(L_pm[, order], decreasing = TRUE)
  L_pm_sorted <- L_pm[sorted_indices, ]
  
  # Convert to a data frame for ggplot
  df <- as.data.frame(L_pm_sorted)
  df$Gene <- factor(1:nrow(L_pm), levels = 1:nrow(L_pm))  # Gene IDs after sorting
  df_long <- reshape2::melt(df, id.vars = "Gene", variable.name = "Factor", value.name = "Intensity")
  ggplot(df_long, aes(x = Gene, y = Intensity, fill = Factor)) +
    geom_bar(stat = "identity") +
    scale_fill_brewer(palette = "Set3") +
    theme_minimal() +
    labs(
      title = paste("Gene Intensity Across Factors (Sorted by Factor", order, ")"),
      x = "Genes (Sorted by Intensity)",
      y = "Intensity",
      fill = "Factor"
    ) +
    theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          panel.grid.major.x = element_blank())
}
plot_L_pm(L_gp_norm, order = 2)
```

```{r fig.width=6, fig.height=6}
loading_data_ebmf <- data.frame(
  L1 = f.mod$L_pm[,1],
  L2 = f.mod$L_pm[,2],
  Gene = Iyer_expression$`Gene name`
)
ggplot(loading_data_ebmf, aes(x = L1, y = L2)) +
  geom_point(color = "blue", size = 0.5) +
  # geom_text(aes(label = Gene), hjust = 0.5, vjust = -0.5, size = 0.5, check_overlap = TRUE) +
  labs(
    title = "EBMF Loading Plot",
    x = "Loading 1",
    y = "Loading 2"
  ) +
  theme_minimal()
```

